---
title: "Goal 1"
author: "Shelby Tisdale"
format: html
editor: visual
---

# Load Data and Packages

```{r}
#| label: load-packages

library(tidyverse)
library(data.table)
library(bit64)
library(janitor)
library(gt)
library(gtsummary)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(tidymodels)
library(bayesrules)
library(broom.mixed)
```

```{r}
#| label: load-full-data
all_goals <- read_csv("data/sipp_all_goals.csv")
```

# Goal 1

**Goal 1:** Provide assistance to needy families so that children can be cared for in their own homes or in the homes of relatives.

**Criteria:** This goal is fulfilled for a given household and month if all children in the household have at least one relative in the household.

**Model:** This model will evaluate the impacts of state-level unobligated funds on household-level fulfillment of TANF Goal 1.

## Aggregate Dataset to Household-Year Level

-   Should this be aggregated to the Household-year or Household-month Level?

    -   I'm thinking household year.

-   Should I include poverty status or income as a covariate, given that all individuals in the sample are in poverty? Should I change the sample to include people of different all income levels? My initial thoughts are that since the only people eligible for TANF are families in poverty, this is who we should model for.

    -   What should I control for?
    -   household level demographics - householder

-   look at how unobligated funds change year to year

```{r}

# create race of householder variables and goal1 variable
hh_demogs <- all_goals |>
  filter(is_householder == 1) |>
  filter(in_goal4 == 1) |>
  select(household_id, state, race, fiscal_year) |>
  distinct() |>
  mutate(
    race = factor(race),
    race = fct_collapse(
      race,
      "White" = c("White"),
      "Black" = c("Black", "Black-AIAN", "White-Black"),
      "AIAN" = c("AIAN", "White-AIAN"),
      "Asian" = c("Asian", "White-Asian"),
      "Other" = c("Other", "HP")
    ),
   race = fct_relevel(race, c("White", "Black", "Asian", "AIAN", "Other"))
  ) |>
  distinct() |>
  group_by(household_id, state, fiscal_year) |>
  summarize(race = first(race), .groups = "drop")

goal1 <- all_goals |>
  # selects all households in goal 1 universe
  filter(in_goal1 == 1) |>
  # select relevant variables
  select(household_id, state, fiscal_year, pct_unobligated, pct_basic_assistance, pct_fatherhood_programs, pct_child_welfare, goal1, tanf_amt_received, party) |>
  group_by(household_id, state, goal1, fiscal_year) |>
  # summary variables
  mutate(
    # amount of tanf received during year
    tanf_amt_received = sum(tanf_amt_received)/100) |>
  ungroup() |>
  distinct() |>
  left_join(hh_demogs, by = c("household_id", "state", "fiscal_year")) |>
  mutate(pct_unobligated = pct_unobligated*100,
         pct_basic_assistance = pct_basic_assistance*100,
         pct_fatherhood_programs = pct_fatherhood_programs*100,
         pct_child_welfare = pct_child_welfare*100,
         is_covid = if_else(fiscal_year >= 2020, 1, 0)) |>
  filter(!is.na(goal1)) |>
  group_by(household_id, state, fiscal_year, pct_unobligated, pct_basic_assistance, pct_fatherhood_programs, pct_child_welfare, tanf_amt_received, party, race, is_covid) |>
  summarize(goal1 = max(goal1, na.rm = TRUE), .groups = "drop")

saveRDS(goal1, "data/goal1.rds")
```

## Table 1

## EDA

## Modeling

**Note:** Aggregate to household level?

Let $Y_{ij1}$ denote whether or not household $i$ in state $j$ met the criteria for Goal 1.

$$ Y_{ij1} = \begin{cases} 1 & \text{yes} \\ 0 & \text{no} \end{cases} $$

```{r}
set.seed(812)

goal1_split <- initial_split(goal1)
goal1_train <- training(goal1_split)
goal1_test <- testing(goal1_split)

# tune prior
goal_1_model_1_prior <- stan_glmer(
  goal1 ~ pct_unobligated + pct_basic_assistance + pct_fatherhood_programs + pct_child_welfare + tanf_amt_received + is_covid +  party + race + (1 | household_id) + (1 | state),
  data = goal1_train, family = binomial,
  # https://www.aecf.org/blog/child-welfare-and-foster-care-statistics#:~:text=Children%20Entering%20Foster%20Care,care%20by%20race%20and%20ethnicity
  # 3/1000 children in foster care
  prior_intercept = normal(-5.81, 1.5, autoscale = TRUE),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_covariance = decov(reg = 1, conc = 1, shape = 1, scale = 1),
  chain = 10, iter = 10000, seed = 12345,
  prior_PD = TRUE)


# summarize prior
prior_summary(goal_1_model_1_prior)
tidy(goal_1_model_1_prior)

# model results
goal_1_model_1 <- update(goal_1_model_1_prior, prior_PD = FALSE)
goal1_model1_summary <- tidy(goal_1_model_1, effects = "fixed", conf.int = TRUE, conf.level = .80)
goal1_model1_summary_or <- goal1_model1_summary |>
  mutate(odd_ratio = exp(estimate),
         conf.low_or = exp(conf.low),
         conf.high_or = exp(conf.high),
         odd_ratio = format(odd_ratio, scientific = F),
         conf.low_or = format(conf.low_or, scientific = F),
         conf.high_or = format(conf.high_or, scientific = F)
         ) |>
  select(term, odd_ratio, conf.low_or, conf.high_or)

saveRDS(goal_1_model_1, "data/goal1_model1.rds")
# goal_1_model_1 <- readRDS("data/goal1_model1.rds")
params <- get_variables(goal_1_model_1)

# model diagnostics
for (i in 1:length(params)) {
  print(mcmc_trace(goal_1_model_1, pars = params[[i]]))
}

for (i in 1:length(params)) {
  print(mcmc_dens_overlay(goal_1_model_1, pars = params[[i]]))
}

for (i in 1:length(params)) {
  print(mcmc_acf(goal_1_model_1, pars = params[[i]]))
}

neff_ratio(goal_1_model_1)
rhat(goal_1_model_1)
prop_goal3 <- function(x){mean(x == 1)}
pp_check(goal_1_model_1, nreps = 100, plotfun = "stat", stat = "prop_goal3")

# classify with testing data
# posterior predictive models
goal1_pred1 <- posterior_predict(goal_1_model_1, newdata = drop_na(goal1_test))
goal1_classifications <- goal1_test |>
  drop_na() |>
  mutate(goal1_prob = colMeans(goal1_pred1),
         goal1_class_1 = as.numeric(goal1_prob >= 0.5)) |>
  select(goal1_prob, goal1_class_1, goal1)

# confusion matrix
goal1_classifications |>
  tabyl(goal1, goal1_class_1) |>
  adorn_totals(c("row", "col"))

goal1_class_summary <- classification_summary(
  model = goal_1_model_1, data = goal1_test |> drop_na(), cutoff = 0.5)

# specificity: of the households that do not meet goal1, model correctly classifies __ percent.
# sensitivity: of the households that do meet goal4, model correctly classifies __ percent.
# a total __ percent of households are correctly classified by this model.
goal1_class_summary$confusion_matrix
goal1_class_summary$accuracy_rates
```

```{r}
# no state-level grouping
# using sample

set.seed(812)

goal1_sample_hh <- goal1 |>
  select(household_id) |>
  sample_n(500)

goal1_sample <- goal1 |>
  filter(household_id %in% goal1_sample_hh$household_id)

# tune prior
goal_1_model_2_prior <- stan_glmer(
  goal1 ~ pct_unobligated + pct_basic_assistance + pct_fatherhood_programs + pct_child_welfare + tanf_amt_received + is_covid +  party + race + (1 | household_id),
  data = goal1, family = binomial,
  # https://www.childtrends.org/publications/data-on-families-with-low-incomes-across-america-can-inform-two-generation-approaches
  # about 51%  of low-income families are two-parent families
  prior_intercept = normal(0.017, 1.5, autoscale = TRUE),
  prior = normal(0, 2.5, autoscale = TRUE),
  prior_covariance = decov(reg = 1, conc = 1, shape = 1, scale = 1),
  chain = 10, iter = 10000, seed = 12345,
  prior_PD = TRUE)


# summarize prior
prior_summary(goal_1_model_2_prior)
tidy(goal_1_model_2_prior)

# model results
goal_1_model_2 <- update(goal_1_model_2_prior, prior_PD = FALSE)
summary(goal_1_model_2)
goal1_model2_summary <- tidy(goal_1_model_2, effects = "fixed", conf.int = TRUE, conf.level = .80)
goal1_model2_summary_or <- goal1_model2_summary |>
  mutate(odd_ratio = exp(estimate),
         conf.low_or = exp(conf.low),
         conf.high_or = exp(conf.high),
         odd_ratio = format(odd_ratio, scientific = F),
         conf.low_or = format(conf.low_or, scientific = F),
         conf.high_or = format(conf.high_or, scientific = F)
         ) |>
  select(term, odd_ratio, conf.low_or, conf.high_or)

saveRDS(goal_1_model_2, "data/goal1_model2.rds")
goal_1_model_2 <- readRDS("data/goal_1_model2.rds")
params <- get_variables(goal_1_model_2)

# model diagnostics
for (i in 1:length(params)) {
  print(mcmc_trace(goal_1_model_2, pars = params[[i]]))
}

for (i in 1:length(params)) {
  print(mcmc_dens_overlay(goal_1_model_2, pars = params[[i]]))
}

for (i in 1:length(params)) {
  print(mcmc_acf(goal_1_model_2, pars = params[[i]]))
}

neff_ratio(goal_1_model_2)
rhat(goal_1_model_2)
prop_goal3 <- function(x){mean(x == 1)}
pp_check(goal_1_model_2, nreps = 100, plotfun = "stat", stat = "prop_goal3")

# classify with testing data
# posterior predictive models
goal1_pred2 <- posterior_predict(goal_1_model_2, newdata = drop_na(goal1_test))
goal1_classifications <- goal1_test |>
  drop_na() |>
  mutate(goal1_prob = colMeans(goal1_pred2),
         goal1_class_2 = as.numeric(goal1_prob >= 0.5)) |>
  select(goal1_prob, goal1_class_1, goal1)

# confusion matrix
goal1_classifications |>
  tabyl(goal1, goal1_class_2) |>
  adorn_totals(c("row", "col"))

goal1_class_summary2 <- classification_summary(
  model = goal_1_model_2, data = goal1_test |> drop_na(), cutoff = 0.5)

# specificity: of the households that do not meet goal1, model correctly classifies __ percent.
# sensitivity: of the households that do meet goal4, model correctly classifies __ percent.
# a total __ percent of households are correctly classified by this model.
goal1_class_summary2$confusion_matrix
goal1_class_summary2$accuracy_rates

sum1 <- summary(goal1_fit) |> as.data.frame()

```

\
\# AGGREGATE DATA BY HOUSEHOLD
