---
title: "Creation of Dataset for Hierarchical Model"
author: "Shelby Tisdale"
format: html
---

# Load Data and Packages

```{r}
#| label: load-packages
library(tidyverse)
library(data.table)
library(bit64)
library(janitor)
```

```{r}
#| label: load-tanf-data
#| message: false
tanf <- read_csv("data/total_expenditures_2015_2022.csv")
```

```{r}
#| label: state-year-tanf-data

# creates dataset of relevant state-year-level tanf predictors 2018-2022
tanf_state_year <- tanf |>
  # filter for relevant years
  filter(fiscal_year %in% c(2017:2021)) |>
  # create variables for percent unobligated funds, percent spent on basic
  # assistance, and percent spent on program management
  mutate(pct_unobligated = unobligated_balance/(funds_awarded + carryover),
         pct_basic_assistance = basic_assistance/total_expenditures,
         pct_program_mgmt = program_mgmt/total_expenditures) |>
  # select relevant variables
  select(state, fiscal_year, pct_unobligated, pct_program_mgmt,
         pct_basic_assistance)
```

```{r}
#| label: load-sipp-data

# create function for loading data
load_sipp <- function(directory) {
  ds <- c(directory)
  sipp <- fread(ds, sep = "|", select = c(
    # ID VARS
   "SHHADID", "SPANEL", "SSUID", "SWAVE", "PNUM", "MONTHCODE", "GHLFSAM",
   "EHRESIDENCID", "EINTTYPE", "ERESIDENCEID","RIN_UNIV",
   "THHLDSTATUS", "TST_INTV", "EPNPAR1", "EPNPAR2", "EPNCOHAB", "EPNSPOUSE",
   "RPNCHILD1", "RPNCHILD2", "RPNCHILD3", "RPNCHILD4", "RPNCHILD5", "RPNCHILD6",
   "RPNCHILD7", "RPNCHILD8", "RPNCHILD9", "RPNCHILD10", "RPNCHILD11",
   "RPNCHILD12",
    # WEIGHTS AND SAMPLING
   "GVARSTR", "WPFINWGT",
   # OTHER
   "EANYKID", "EBDAD", "EBMOM", "ECLTH_SR1YN", "EJB1_BSLRYB", "EJB1_SCRNR",
   "EMPF", "EMS_EHC", "ENJ_LKWRK", "ENJFLAG", "EPAR_SCRNR", "EPAR1TYP",
   "EPAR2TYP", "ERELRPE", "ETANF_BMONTH", "ETANF_COVTYP", "ETANF_EMONTH",
   "ETANF_OWN", "ETANF_PTCS", "EWELAC_MNYN", "EWELACTV1", "EWELACTV2_1",
   "EWELACTV2_2", "EWELACTV2_3", "EWELACTV4", "EXMAR", "RCB_UNION", "RCHTYP1",
   "RCHTYP2", "RCHTYP3", "RCHTYP4","RCHTYP5", "RCHTYP6", "RCHTYP7","RCHTYP8",
   "RCHTYP9", "RCHTYP10", "RCHTYP11", "RCHTYP12", "RSNAP_YRYN", "RTANF_CFLG",
   "RTANF_LCYR", "RTANF_MNYN", "RGA_YRYN", "RTANF_YRYN", "RWIC_YRYN", "TAGE",
   "TAGE_FB", "EUC1MNYN",
   "TCEB", "TFCYINCPOV", "TFINCPOV", "THINCPOV", "TNUM_CBU", "TNUMKIDS",
   "TPTRNINC", "TTANF_AMT", "TTANF_BRSN", "TTANF_ERSN", "TYEAR_FB", "ERP",
   "RFAMNUM", "RFPERSONS",
   # DEMOGRAPHICS
   "EEDUC", "EORIGIN", "TRACE", "ESEX", "RAFEVER", "ECITIZEN",
   # RELATIONSHIP DATA
   "RREL1", "RREL2", "RREL3", "RREL4", "RREL5", "RREL6", "RREL7", "RREL8",
   "RREL9", "RREL10", "RREL11", "RREL12", "RREL13", "RREL14", "RREL15",
   "RREL16", "RREL17", "RREL18", "RREL19", "RREL20", "RREL21", "RREL22",
   "RREL23", "RREL24", "RREL25", "RREL26", "RREL27", "RREL28", "RREL29",
   "RREL30", "RREL_PNUM1", "RREL_PNUM2", "RREL_PNUM3", "RREL_PNUM4",
   "RREL_PNUM5", "RREL_PNUM6", "RREL_PNUM7", "RREL_PNUM8", "RREL_PNUM9",
   "RREL_PNUM10", "RREL_PNUM11", "RREL_PNUM12", "RREL_PNUM13", "RREL_PNUM14",
   "RREL_PNUM15", "RREL_PNUM16", "RREL_PNUM17", "RREL_PNUM18", "RREL_PNUM19",
   "RREL_PNUM20", "RREL_PNUM21", "RREL_PNUM22", "RREL_PNUM23", "RREL_PNUM24",
   "RREL_PNUM25", "RREL_PNUM26", "RREL_PNUM27", "RREL_PNUM28", "RREL_PNUM29",
   "RREL_PNUM30",
   # CHILD BIRTH YEARS
   "TCBYR_1", "TCBYR_2", "TCBYR_3", "TCBYR_4", "TCBYR_5", "TCBYR_6", "TCBYR_7"
  ))
  names(sipp) <- toupper(names(sipp))
  return(sipp)
}
# load survey data for each year
sipp2022 <- load_sipp("data/sipp/pu2022.csv") |>
  mutate(year = 2021)
sipp2021 <- load_sipp("data/sipp/pu2021.csv") |>
  mutate(year = 2020)
sipp2020 <- load_sipp("data/sipp/pu2020.csv") |>
  mutate(year = 2019)
sipp2019 <- load_sipp("data/sipp/pu2019.csv") |>
  mutate(year = 2018)
sipp2018 <- load_sipp("data/sipp/pu2018.csv") |>
  mutate(year = 2017)

# combine data frames
sipp <- rbind.data.frame(sipp2022, sipp2021, sipp2020, sipp2019, sipp2018) |>
  clean_names()

# download csv
# write.csv(sipp, "sipp.csv")
```

```{r}
#| label: recode-states

sipp <- sipp |>
  mutate(state = case_match(
    tst_intv,
    1 ~ "ALABAMA",
    2 ~ "ALASKA", 
    4 ~ "ARIZONA",
    5 ~ "ARKANSAS",
    6 ~ "CALIFORNIA",
    8 ~ "COLORADO",
    9 ~ "CONNECTICUT",
    10 ~ "DELAWARE",
    11 ~ "DIST.OF COLUMBIA",
    12 ~ "FLORIDA",
    13 ~ "GEORGIA",
    15 ~ "HAWAII",
    16 ~ "IDAHO",
    17 ~ "ILLINOIS",
    18 ~ "INDIANA",
    19 ~ "IOWA",
    20 ~ "KANSAS",
    21 ~ "KENTUCKY",
    22 ~ "LOUISIANA",
    23 ~ "MAINE",
    24 ~ "MARYLAND",
    25 ~ "MASSACHUSETTS",
    26 ~ "MICHIGAN",
    27 ~ "MINNESOTA",
    28 ~ "MISSISSIPPI",
    29 ~ "MISSOURI",
    30 ~ "MONTANA",
    31 ~ "NEBRASKA",
    32 ~ "NEVADA",
    33 ~ "NEW HAMPSHIRE",
    34 ~ "NEW JERSEY",
    35 ~ "NEW MEXICO",
    36 ~ "NEW YORK",
    37 ~ "NORTH CAROLINA",
    38 ~ "NORTH DAKOTA",
    39 ~ "OHIO",
    40 ~ "OKLAHOMA",
    41 ~ "OREGON",
    42 ~ "PENNSYLVANIA",
    44 ~ "RHODE ISLAND",
    45 ~ "SOUTH CAROLINA",
    46 ~ "SOUTH DAKOTA",
    47 ~ "TENNESSEE",
    48 ~ "TEXAS",
    49 ~ "UTAH",
    50 ~ "VERMONT",
    51 ~ "VIRGINIA",
    53 ~ "WASHINGTON",
    54 ~ "WEST VIRGINIA",
    55 ~ "WISCONSIN",
    56 ~ "WYOMING",
    60 ~ "PUERTO RICO AND ISLAND AREAS",
    61 ~ "FOREIGN COUNTRY",
    .default = as.character(tst_intv)
    )) |>
  filter(!is.na(state))
```

# Filter Data for Relevant Participants

This dataset includes all low-income households that have at least one child in at least one month. Low-income households will be defined by looking at data on TANF eligibility by income and state. Households with TANF-eligible incomes will be included in the sample.

```{r}
#| label: find-participants
#| warning: false

# find all households that have at least one child in any month
households_with_children <- sipp |>
  # sample unit id is originally sampled household
  # this will be used as an ID variable for household
  # look at each household for each month it is in sample
  group_by(ssuid, year, monthcode) |>
  # count number of children in each household
  summarize(hh_children_count = sum(tage < 18, na.rm = TRUE), .groups = "drop") |>
  # remove households without children
  filter(hh_children_count != 0) |>
  # select distinct households
  select(ssuid) |>
  distinct()

# excluding sample to families below poverty line
# only families in deep poverty are eligible to receive TANF
# find households who are below poverty line for at least one month in sample
households_poverty <- sipp |>
  group_by(ssuid, year, monthcode) |>
  summarize(poverty = if_else(thincpov <= 1, 1, 0), .groups = "drop") |>
  filter(poverty == 1) |>
  select(ssuid) |>
  distinct()

# dataset of low-income households with children
sipp_participants <- sipp |>
  filter(ssuid %in% households_with_children$ssuid &
      ssuid %in% households_poverty$ssuid)
```

# Creating Variable for Achievement of TANF Goal 1

**Goal 1:** Provide assistance to needy families so that children can be cared for in their own homes or in the homes of relatives.

**Criteria:** This goal is fulfilled for a given household and month if all children in the household have at least one relative in the household.

```{r}
#| label: tanf-goal-1

# 1. Filter for all children in dataset
# 2. Check whether child has parents in household
# 3. Check child's relationship with householder.
# 4. For children who are not related to householder,
# check all household relationships for that child

# If child has relative or parent in household, goal is fulfilled for that child.
# If all children in the household have a parent or relative in the household,
# then goal is fulfilled for household

children <- sipp_participants |>
  # filters for children
  filter(tage < 18) |>
  # select id variables
  select(ssuid, year, monthcode, pnum, erelrpe, epnpar1, epnpar2,
         starts_with("rrel")) |>
    mutate(goal1 = case_when(
    # child who has at least one parent in household
    !is.na(epnpar1) | !is.na(epnpar2) ~ 1,
    # child is householder with relatives
    # applies to a few older children
    is.na(epnpar1) & is.na(epnpar2) & erelrpe == 1 ~ 1,
    # foster child
    is.na(epnpar1) & is.na(epnpar2) & erelrpe == 15 ~ 0,
    # grandchild
    is.na(epnpar1) & is.na(epnpar2) & erelrpe == 8 ~ 1,
    # sibling
    is.na(epnpar1) & is.na(epnpar2) & erelrpe == 10 ~ 1,
    # niece / nephew
    is.na(epnpar1) & is.na(epnpar2) & erelrpe == 13 ~ 1,
    # other relative
    is.na(epnpar1) & is.na(epnpar2) & erelrpe == 14 ~ 1,
    # indicator for fringe cases
    .default = 2
  ))


# create lookup table for children and household relationships
# for fringe cases (no parent in household, housholder is not relative)
children_lookup <- children |>
  # filter for fringe cases
  filter(goal1 == 2) |>
  rename(child_pnum = pnum) |>
  # child person number and sample unit id (id vars)
  # relationship variables
  select(child_pnum, ssuid, monthcode, year, erelrpe,
         starts_with("rrel"), goal1) |>
  # each row is a household member-child combination for a month, for
  # person numbers of all  members of the household
  pivot_longer(cols = starts_with("rrel_pnum"),
               names_to = "hh_member_number",
               values_to = "person_num_rel") |>
  filter(!is.na(person_num_rel)) |>
  # each row is a household member-child combination for a month, for the
  # person number and relationship to all members of household
  pivot_longer(cols = starts_with("rrel"),
               names_to = "hh_member_number2",
               values_to = "relationship") |>
  filter(!is.na(relationship)) |>
  # match household member numbers for relationship and person number variables
  mutate(hh_member_number = parse_number(hh_member_number),
         hh_member_number2 = parse_number(hh_member_number2)) |>
  filter(hh_member_number == hh_member_number2) |>
  # remove relationship between child and self
  filter(relationship != 99) |>
  # select relevant variables for lookup
  select(child_pnum, ssuid, year, monthcode, relationship, goal1) |>
  # indicator for whether person is a relative of the child
  mutate(goal1_lookup = if_else(
    relationship %in% c(5, 7, 8, 9, 10, 12, 16, 17), 1, 0)) |>
  group_by(year, monthcode, ssuid, child_pnum) |>
  # if average of relative indicators is greater than 0, child lives with at
  # least one relative
  summarize(goal1_avg = mean(goal1, na.rm = TRUE), .groups = "drop") |>
  mutate(goal1_lookup = if_else(goal1_avg == 0, 0, 1)) |>
  select(-c(goal1_avg)) 

# match children dataset to lookup table to update goal fulfilled for fringe cases
children <- children |>
  left_join(children_lookup,
            by = join_by("pnum" == "child_pnum", year, monthcode, ssuid)) |>
  select(year, monthcode, ssuid, pnum, goal1, goal1_lookup) |>
  mutate(goal1 = if_else(goal1 == 2, goal1_lookup, goal1))

# evaluates whether tanf goal 1 is achieved for a household in a given year and month
hh_goal1 <- children |>
  # group by year, monthj, and household
  group_by(year, monthcode, ssuid) |>
  # determine average (over all children in household in that month)
  # achievement of goal 1
  summarise(avg_goal1 = mean(goal1, na.rm = TRUE), .groups = "drop") |>
  # if goal1 = 0 for any child in household, goal1 is not achieved
  mutate(hh_month_goal1 = if_else(avg_goal1 != 1, 0, 1)) |>
  select(-avg_goal1)

filter(hh_goal1, is.na(hh_month_goal1))d
```

# **Creating Variables for Achievement of TANF Goal 2**

**Goal 2:** End the dependence of needy parents on government benefits by promoting job preparation, work, and marriage.

**Participant Households:** All households who received government benefits in the earliest year of their participation in the study.

**Criteria:** Goal 2 is fulfilled for a participant household if household members are no longer receiving government benefits by the end of the reference frame. Variables related to job preparation, work, and marriage will be used as predictors.

**Note:** Year will not be a level in this model

```{r}
#| label: tanf-goal-2


# find households receiving benefits at beginning of reference period
govt_benefits_begin <- sipp_participants |>
  group_by(ssuid) |>
  # find first year of reference period for household
  filter(year == min(year)) |>
  ungroup() |>
  # detect whether person received benefits
  mutate(govt_benefits_begin = if_else(
    rga_yryn == 1 | rtanf_yryn == 1 | rsnap_yryn == 1 | rwic_yryn == 1, 1, 0
  )) |>
  # summarize benefit receipt per household
  group_by(year, ssuid) |>
  summarize(govt_benefits_begin = mean(govt_benefits_begin), .groups = "drop") |>
  mutate(govt_benefits_begin = if_else(govt_benefits_begin > 0, 1, 0)) |>
  # save variable for first year of reference period
  rename(begin_year = year)

# find whether households received benefits at end of reference period
govt_benefits_end <- sipp_participants |>
  group_by(ssuid) |>
  # find last year of reference period for household
  filter(year == max(year)) |>
  ungroup() |>
  # detect whethter individual received benefits
  mutate(govt_benefits_end = if_else(
    rga_yryn == 1 | rtanf_yryn == 1 | rsnap_yryn == 1 | rwic_yryn == 1, 1, 0
  )) |>
  # summarize benefit receipt per houeshold
  group_by(year, ssuid) |>
  summarize(govt_benefits_end = mean(govt_benefits_end), .groups = "drop") |>
  mutate(govt_benefits_end = if_else(govt_benefits_end > 0, 1, 0)) |>
  # save variable for last year of reference period
  rename(end_year = year)

# join govt benefits datasets
goal2 <- inner_join(govt_benefits_begin, govt_benefits_end, by = join_by("ssuid")) |>
  # filter for households that are in dataset for more than one year
  # and are receiving government benefits in first year
  filter(begin_year != end_year & govt_benefits_begin == 1) |>
  # create variable for whether the household is still receiving benefits at the end of the reference period
  # 1: goal fulfilled, family is no longer receiving goverment benefits
  # 0: goal not fulfilled
  mutate(goal2 = if_else(govt_benefits_end < govt_benefits_begin, 1, 0)) |>
  mutate(goal2_participant = 1)

# create dataset with response and relevant predictors for tanf goal 2
# at person-month level
person_goal2 <- sipp_participants |>
  # create variable for whether individual received job training in month
  mutate(received_job_training = case_when(
    ewelactv2_1 == 1 ~ 1,
    ewelac_mnyn == 1 ~ 1, 
    ewelactv2_2 == 1 ~ 1,
    ewelactv2_3 == 1 ~ 1,
    ewelactv1 == 1 ~ 1,
    ewelactv4 == 1 ~ 1,
    .default = 0
  )) |>
  # whether householder is married in month
  mutate(hh_married_month = case_when(
    (erelrpe == 1 | erelrpe == 2) & (ems_ehc == 1 | ems_ehc == 2) ~ 1,
    (erelrpe == 1 | erelrpe == 2) & !(ems_ehc == 1 | ems_ehc == 2) ~ 0,
    .default = NA)
  ) |>
  left_join(goal2, join_by(ssuid)) |>
  mutate(goal2_participant = if_else(is.na(goal2_participant), 0, 1)) |>
  select(ssuid, pnum, year, monthcode, received_job_training, hh_married_month, goal2) |>
  distinct()
```

# **Creating Variable for Achievement of TANF Goal 3**

**Goal 3:** Prevent and reduce the incidence of out-of-wedlock pregnancies

**Note:** Evaluate at person-level instead of household-level

**Participants:** Women age 15+, for all years in which they are unmarried

**Criteria:** Goal 3 is fulfilled if number of children of an unmarried individual does not increase during reference period (will not catch all cases due to length of pregnancies, abortions, and miscarriages, but should catch most)

```{r}
#| label: tanf-goal-3

goal_3_women <- sipp_participants |>
  # filters for women between ages 15 and 50 who are not married
  filter(tage >= 15 & tage <= 50 & ems_ehc != 1 & ems_ehc != 2 & esex == 2) |>
  # select id variables and variables for birth years of children
  select(ssuid, monthcode, year, pnum, tceb, tcbyr_1:tcbyr_7) |>
  # turn year and month into date object
  mutate(month = as.character(monthcode)) |>
  mutate(month = if_else(nchar(month) == 1, paste("0", month, sep = ""),
                         month)) |>
  mutate(year_month_day = ymd(paste(year, month, "01", sep = "-"))) |>
  group_by(ssuid, pnum) |>
  # find minimum and maximum date that each woman is unmarried and in sample
  mutate(min_date = min(year_month_day), max_date = max(year_month_day)) |>
  ungroup() |>
  # pivot data so that each row is the birth year of one of the children of each woman
  pivot_longer(
    cols = starts_with("tcbyr"),
    names_to = "child_indicator",
    values_to = "birth_year"
  ) |>
  # remove observations with no birth year
  filter(!is.na(birth_year)) |>
  # create indicator for whether woman gave birth during time unmarried and in sample
  mutate(had_child_unmarried = if_else(
    birth_year > year(min_date) & birth_year < year(max_date), 1, 0)) |>
  # creates data set of women who had children out-of-wedlock during time in sample
  filter(had_child_unmarried == 1) |>
  select(ssuid, pnum, had_child_unmarried) |>
  distinct()

# checking goal3 indicator with contextual data
# indicator makes sense
goal3_test <- sipp_participants |>
  select(ssuid, pnum, monthcode, year, epar_scrnr, tceb, tage, rcb_union,
         ems_ehc, epncohab, rpnchild1:rpnchild12) |>
  right_join(goal_3_women, by = join_by(ssuid, pnum)) |>
  arrange(ssuid, pnum, year, monthcode)

# goal 3 participants with indicator for whether goal 3 is fulfilled
# disregards year and month levels
goal3 <- sipp_participants |>
  filter(tage >= 15 & tage <= 50 & ems_ehc != 1 & ems_ehc != 2 & esex == 2) |>
  select(ssuid, pnum) |>
  left_join(goal_3_women, by = join_by(ssuid, pnum)) |>
  mutate(goal3 = if_else(is.na(had_child_unmarried), 1, 0)) |>
  select(-had_child_unmarried) |>
  distinct()
```

# **Creating Variables for Achievement of TANF Goal 4**

**Goal:** Encourage the formation and maintenance of two-parent families.

**Participants:** Households with children and at least one parent

**Criteria:** one parent family is two-parent family at end of reference period, two-parent family is still two-parent family

```{r}
#| label: tanf-goal-4

# finds number of parents that each child has for each year and month
child_parent_count <- sipp_participants |>
  # filters participants for children only
  mutate(is_child = if_else(tage < 18, 1, 0)) |>
  filter(is_child == 1) |>
  # select relevant variables
  select(ssuid, pnum, tage, year, monthcode, epnpar1, epnpar2, epar1typ,
         epar2typ) |>
  # create variable for the number of parents a child has
  mutate(num_parents = case_when(
    is.na(epnpar1) & is.na(epnpar2) ~ 0,
    is.na(epnpar1) | is.na(epnpar2) ~ 1,
    !is.na(epnpar1) & !is.na(epnpar2) ~ 2
  ))

# finds the number of parents recorded for a child at the end of the reference period
child_parent_count_end <- child_parent_count |>
  group_by(ssuid, pnum) |>
  filter(year == max(year)) |> 
  ungroup() |>
  select(ssuid, pnum, year, num_parents) |>
  rename(num_parents_end = num_parents,
         max_year = year) |>
  distinct()

# finds the number of parents recorded for a child at the beginning of the reference period
child_parent_count_start <- child_parent_count |>
  group_by(ssuid, pnum) |>
  filter(year == min(year)) |> 
  ungroup() |>
  select(ssuid, pnum, year, num_parents) |>
  rename(num_parents_start = num_parents,
         min_year = year) |>
  distinct()

# creates indicator for whether a goal is achieved
goal4 <- inner_join(child_parent_count_start, child_parent_count_end,
                    by = join_by("ssuid", "pnum")) |>
  # removes children who don't have parents to begin with
  filter(num_parents_start != 0) |>
  mutate(goal4 = if_else(num_parents_end == 2, 1, 0))

# evaluates whether goal 4 is achieved for a household
hh_goal4 <- goal4 |>
  # group by household
  group_by(ssuid) |>
  # determine average (over all children in household in that month)
  # achievement of goal 1
  summarise(avg_goal4 = mean(goal4, na.rm = TRUE), .groups = "drop") |>
  # if goal1 = 0 for any child in household, goal1 is not achieved
  mutate(hh_goal4 = if_else(avg_goal4 != 1, 0, 1))

```

```{r}
#| label: data-all-goals

sipp_all_goals <- sipp_participants |>
  # select necessary variables
  select(ssuid, pnum, monthcode, year, tage, state, eeduc, eorigin, trace, esex, ecitizen, tceb,
  # tanf variables
  etanf_bmonth, rtanf_lcyr, etanf_emonth, etanf_covtyp, etanf_own, rtanf_cflg,
  rtanf_mnyn, rtanf_yryn, ttanf_amt) |>
  # clean up variables 
  mutate(
    education = case_match(
      eeduc,
      31 ~ "Less than 1st grade",
      32 ~ "1st, 2nd, 3rd or 4th grade",
      33 ~ "5th or 6th grade",
      34 ~ "7th or 8th grade",
      35 ~ "9th grade",
      36 ~ "10th grade",
      37 ~ "11th grade",
      38 ~ "12th grade, no diploma",
      39 ~ "High School Graduate",
      40 ~ "Some college credit, but less than 1 year",
      41 ~ "1 or more years of college, no degree",
      42 ~ "Associate's degree",
      43 ~ "Bachelor's degree",
      44 ~ "Master's degree",
      45 ~ "Professional School degree",
      46 ~ "Doctorate degree",
      .default = as.character(eeduc)
    ),
    ethnicity = if_else(eorigin == 1, "Hispanic", "Not Hispanic"),
    sex = if_else(esex == 1, "Male", "Female"),
    race = case_match(
      trace,
      1 ~ "White",
      2 ~ "Black",
      3 ~ "AIAN",
      4 ~ "Asian",
      5 ~ "HP",
      6 ~ "White-Black",
      7 ~ "White-AIAN",
      8 ~ "White-Asian",
      9 ~ "Black-AIAN",
      10 ~ "Other"
    ),
    citizenship = if_else(ecitizen == 1, 1, 0)
  ) |>
  select(-c(ecitizen, trace, eeduc, esex, eorigin)) |>
  # create indicators for whether this person-month is included in the sample for a certain goal
  mutate(
    # all households are in goal 1
    in_goal1 = if_else(ssuid %in% hh_goal1$ssuid, 1, 0),
    # selected households for goal2
    in_goal2 = if_else(ssuid %in% goal2$ssuid, 1, 0),
    # individuals in goal 3
    in_goal3 = if_else(ssuid %in% goal3$ssuid, 1, 0),
    # all households are in goal 4
    in_goal4 = if_else(ssuid %in% hh_goal4$ssuid, 1, 0)
    ) |>
  # join goal 1 achievement data
  left_join(hh_goal1, by = join_by(year, monthcode, ssuid)) |>
  # join goal 2 achievement data
  left_join(person_goal2, by = join_by("ssuid", "pnum", "monthcode", "year")) |>
  # join goal 3 achievement data
  left_join(goal3, by = join_by("ssuid", "pnum")) |>
  # join goal 4 achievement data
  left_join(hh_goal4, by = join_by("ssuid")) |>
  select(-avg_goal4) |>
  rename(goal1 = hh_month_goal1,
         goal4 = hh_goal4) |>
  # join to tanf data
  left_join(tanf_state_year, by = join_by("state", "year" == "fiscal_year")) |>
  # select variables in desired order
  select(
  # id variables
  ssuid, pnum, monthcode, year,
  # demographics
  tage, state, education, ethnicity, race, sex, citizenship, tceb,
  # goal fulfillment indicators
  in_goal1, in_goal2, in_goal3, in_goal4, goal1, goal2, goal3, goal4,
  # state-level predictors
  pct_unobligated, pct_program_mgmt, pct_basic_assistance,
  # individual predictors
  received_job_training, hh_married_month,
  # tanf variables
  etanf_bmonth, rtanf_lcyr, etanf_emonth, etanf_covtyp, etanf_own,
  rtanf_cflg, rtanf_mnyn, rtanf_yryn, ttanf_amt
  ) |>
  # variable for party of governor in that year
  mutate(party = case_when(
    state == "ALABAMA" ~ "R",
    state == "ALASKA" & year > 2018 ~ "R",
    state == "ALASKA" & year <= 2018 ~ "I",
    state == "ARIZONA" ~ "R",
    state == "ARKANSAS" ~ "R",
    state == "CALIFORNIA" ~ "D",
    state == "COLORADO" ~ "D",
    state == "CONNECTICUT" ~ "D",
    state == "DELAWARE" ~ "D",
    state == "DIST.OF COLUMBIA" ~ "D",
    state == "FLORIDA" ~ "R",
    state == "GEORGIA"~ "R",
    state == "HAWAII" ~ "D",
    state == "IDAHO" ~ "R",
    state == "ILLINOIS" & year <= 2018 ~ "R",
    state == "ILLINOIS" & year > 2018 ~ "D",
    state == "INDIANA" ~ "R",
    state == "IOWA" ~ "R",
    state == "KANSAS" & year <= 2018 ~ "R",
    state == "KANSAS" & year > 2018 ~ "D",
    state == "KENTUCKY" & year <= 2018 ~ "R",
    state == "KENTUCKY" & year > 2018 ~ "D",
    state == "LOUISIANA" ~ "D",
    state == "MAINE" & year <= "2018" ~ "R",
    state == "MAINE" & year > "2018" ~ "D",
    state == "MARYLAND" ~ "R",
    state == "MASSACHUSETTS" ~ "R",
    state == "MICHIGAN" & year <= 2018 ~ "R",
    state == "MICHIGAN" & year > 2018 ~ "D",
    state == "MINNESOTA" ~ "D",
    state == "MISSISSIPPI" ~ "R",
    state == "MISSOURI" ~ "R",
    state == "MONTANA" & year < 2021 ~ "D",
    state == "MONTANA" & year >= 2021 ~ "R",
    state == "NEBRASKA" ~ "R",
    state == "NEVADA" & year <= 2018 ~ "R",
    state == "NEVADA" & year > 2018 ~ "D",
    state == "NEW HAMPSHIRE" ~ "R",
    state == "NEW JERSEY" & year <= 2017 ~ "R",
    state == "NEW JERSEY"& year > 2017 ~ "D",
    state == "NEW MEXICO" & year <= 2018 ~ "R",
    state == "NEW MEXICO" & year > 2018 ~ "D",
    state == "NEW YORK" ~ "D",
    state == "NORTH CAROLINA" ~ "D",
    state == "NORTH DAKOTA" ~ "R",
    state == "OHIO" ~ "R",
    state == "OKLAHOMA" ~ "R",
    state == "OREGON" ~ "D",
    state == "PENNSYLVANIA" ~ "D",
    state == "RHODE ISLAND" ~ "D",
    state == "SOUTH CAROLINA" ~ "R",
    state == "SOUTH DAKOTA" ~ "R",
    state == "TENNESSEE" ~ "R",
    state == "TEXAS" ~ "R",
    state == "UTAH" ~ "R",
    state == "VERMONT" ~ "R",
    state == "VIRGINIA" ~ "D",
    state == "WASHINGTON" ~ "D",
    state == "WEST VIRGINIA" & year < 2020 ~ "D",
    state == "WEST VIRGINIA" & year >= 2020 ~ "R",
    state == "WISCONSIN" & year <= 2018 ~ "R",
    state == "WISCONSIN" & year > 2018 ~ "D",
    state == "WYOMING" ~ "R"
  )) |>
  rename(
    household_id = ssuid,
    month = monthcode,
    age = tage,
    total_children = tceb,
    tanf_begin_month = etanf_bmonth,
    tanf_initial_year = rtanf_lcyr,
    tanf_end_month = etanf_emonth,
    tanf_coverage_type = etanf_covtyp,
    tanf_owner = etanf_own,
    tanf_continue_flag = rtanf_cflg,
    tanf_received_month = rtanf_mnyn,
    tanf_received_year = rtanf_yryn,
    tanf_amt_received = ttanf_amt
  ) |>
  mutate(
    tanf_coverage_type = case_match(
      tanf_coverage_type,
      1 ~ "Only children",
      2 ~ "Only a pregnant woman",
      3 ~ "Both children and adults"
    ),
    tanf_continue_flag = case_match(
      tanf_continue_flag,
      1 ~ "Receipt ended month 12",
      2 ~ "Receipt ended after month 12 and before interview",
      3 ~ "Receipt continued into interview month"
    ),
    tanf_received_month = if_else(tanf_received_month == 1, 1, 0),
    tanf_received_year = if_else(tanf_received_year == 1, 1, 0)
  )

write.csv(sipp_all_goals, "sipp_all_goals.csv")
```

```{r}
sipp_all_goals |>
  filter(!is.na(goal4)) |>
  select(household_id, month, goal4) |>
  distinct() |>
  count(goal4)
```
