---
title: "TANF Visualizations"
author: "Shelby Tisdale"
format: html
---

```{r}
#| label: load-packages-and-data

# load packages
library(tidyverse)
library(extrafont)
library(colorspace)
library(scales)
library(geojsonio)
library(RColorBrewer)
library(broom)
library(sf)

# load data
expenditures <- read_csv("data/total_expenditures_2015_2022.csv")
```

```{r}
#| label: segmented-bar-all-categories
#| fig-height: 14
#| fig-width: 10

# wrangle data for bar plot
expenditures_bar <- expenditures |>
  # pivot all columns except state and fiscal year
  # create category and amount columns
  pivot_longer(
    cols = funds_awarded:unobligated_balance,
    names_to = "category",
    values_to = "amount"
    ) |>
  # calculate mean amount per category and state from 2015-2022
  group_by(category, state) |>
  summarize(average_amount = mean(amount), .groups = "drop") |>
  # assign categories
  mutate(
    category = case_when(
      category %in% c("basic_assistance_no_subsidies", "basic_assistance_subsidies", "other_assistance_foster", "other_assistance_juvenile_justice") ~
        "Direct Cash Assistance",
      category %in% c(
                      "non_assistance_welfare_foster", "child_welfare") ~
        "Child Welfare, Adoption, and Foster Care",
      category %in% c(
                      "non_assistance_juvenile_justice",
                      "children_youth_services") ~
        "Juvenile Justice and Youth Services",
      category %in% c("work_ed", "work_supports", "financial_ed") ~
        "Work, Education, and Training Activities",
      category %in% c("early_ed") ~
        "Early Childcare and Education",
      category %in% c("support__services", "pregnancy_prevention",
                    "fatherhood_programs", "home_visiting") ~
        "Family Support and Formation Services",
      category %in% c("program_mgmt") ~
        "Program Management",
      category %in% c("other", "other_assistance_emergency",
                      "non_assistance_emergency", "eitc_refundable",
                      "short_term_benefits", "non_eitc_refundable") ~
        "Other",
      category == "unobligated_balance" ~ "Unobligated"
  )) |>
  # remove unnecessary values
  filter(!is.na(category))

expenditures_bar |>
  # reverse order of states
  mutate(state = fct_rev(state)) |>
  # move US to top of plot
  mutate(state = fct_relevel(state, "U.S. TOTAL", after = Inf)) |>
  # order categories by mean average amount
  #mutate(category = fct_reorder(category, average_amount, .fun = mean)) |>
  # move Other category to end
  mutate(category = fct_relevel(category, "Other", after = 0)) |>
  mutate(category = fct_relevel(category, "Direct Cash Assistance", after = Inf)) |>
  mutate(category = fct_relevel(category, "Unobligated", after = Inf)) |>
  ggplot(aes(y = fct_rev(state))) +
  geom_col(aes(x = average_amount, fill = category), position = "fill") +
  scale_fill_brewer(palette = "Pastel1")
  
  # plot varplot with state on y axis
  ggplot(expenditures_bar, aes(y = state)) +
  geom_col(aes(x = average_amount, fill = category), position = "fill",
           width = .75) +
  geom_vline(xintercept = c(.25, .5, .75), color = "white", linetype = "dotted", linewidth = 1) + 
  #scale_fill_brewer(palette = "Dark2", direction = -1) +
  # remove extra space on plot, change labels to percents
  scale_x_continuous(expand = c(0,0), labels = percent) +
  theme_minimal() +
  labs(y = NULL, x = "Average Percent of Total Yearly TANF Expenditures",
       title = "Average TANF Expenditures by Category, 2015-2022") +
  theme(legend.position = "top",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_text(color = "black"),
        panel.background = element_rect(fill = "white", color = "white"),
        text = element_text(family = "Lexend"),
        plot.title = element_text(size = 22, face = "bold"),
        plot.subtitle = element_text(size = 16),
        panel.border = element_blank(),
        axis.title.x = element_text(size = 14),
        axis.text.x = element_text(size = 10),
        legend.key.size = unit(.25, 'cm'),
        legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(title = NULL, nrow = 4, reverse = T,
                             override.aes = list(size = 4)))

```

Observations:

-   Mississippi filed most of the misappropriated funds under "TANF Work Program". Mississippi has unusually high expenditures in the "Work, Education, and Training Activities" category, implying that the effect of the welfare scandal on the amount of money actually going to services may have been quite large.

-   Tennessee has a drastically high amount of unobligated funds. This article gives some insight: <https://tennesseelookout.com/2024/01/22/years-after-public-outcry-over-unspent-funds-for-needy-families-tennessee-has-717m-stockpile/>. Similarly for Hawaii, <https://www.tfhawaii.org/wordpress/blog/2023/04/tanf-hoarding-continued/>. While some political organizations and policy research organizations have taken note of the unobligated funds, I have found no research showing the actually impacts of unobligated funds on outcomes for children and families.

-   It's interesting to me that intuitively, unobligated funds would be a bad thing for children and families, and research has shown that percent direct cash assistance definitely has a large impact on child well-being, but there is little relationship between the two (percent cash assistance and percent unobligated)

-   I'm curious why program management funds vary so largely between states. It would make sense for larger states, states with more money to spend would spend more money on program management because there is more to manage, however, this is not necessarily the case.

-   After looking at this, it might be interesting to create a heatmap showing the relationships between these variables; this could be helpful toward understanding what makes the difference in the amount of funds in a state that go toward their intended purposes.

```{r}
#| label: segmented-bar-avg-unobligated
#| fig-height: 14
#| fig-width: 10

# wrangle data for bar plot
expenditures_bar2 <- expenditures |>
  group_by(state) |>
  summarize(mean_unobligated = mean(unobligated_balance/(funds_awarded + carryover +fed_unliquidated))) |>
  ungroup() |>
  mutate(state = fct_rev(state))

expenditures_bar2 |>
  ggplot(aes(y = state)) +
  geom_col(aes(x = mean_unobligated), fill = "#c98443") +
  scale_x_continuous(labels = percent) +
  labs(title = "Average Percent of TANF Funds Unobligated, 2015-2022",
       x = "Percent Unobligated of Available TANF Funds (Awarded Funds + Carryover From Previous Year)",
       y = NULL) +
  theme_minimal() +
  theme(legend.position = "top",
      axis.text.y = element_text(color = "black"),
      panel.background = element_rect(fill = "white", color = "white"),
      text = element_text(family = "Lexend"),
      plot.title = element_text(size = 22, face = "bold"),
      plot.subtitle = element_text(size = 16),
      panel.border = element_blank(),
      axis.title.x = element_text(size = 14),
      axis.text.x = element_text(size = 10),
      legend.key.size = unit(.25, 'cm'),
      legend.text = element_text(size = 8)) +
  guides(fill = guide_legend(title = NULL, nrow = 4, reverse = T,
                             override.aes = list(size = 4))) 
  

```

```{r}
#| label: unobligated-map
#| width: 10
#| height: 5
#| eval: false

# library
library(tidyverse)
library(geojsonio)
library(RColorBrewer)
library(maps)
library(mapdata)
# Download the Hexagones boundaries at geojson format here: https://team.carto.com/u/andrew/tables/andrew.us_states_hexgrid/public/map.

us_hexbin <- st_read("data/us_states_hexgrid.geojson") |>
  mutate(google_name = toupper(str_remove(google_name, " \\(United States\\)"))) |>
  mutate(google_name = if_else(google_name == "DISTRICT OF COLUMBIA", "DIST.OF COLUMBIA", google_name)) |>
  left_join(expenditures_bar2, join_by(google_name == state))


ggplot(us_hexbin) +
  geom_sf(aes(fill = mean_unobligated), color = "white", linewidth = 1) +
  geom_sf_text(aes(label = iso3166_2), color = "black", family = "Lexend") +
  theme_void() +
  scale_fill_distiller(direction = 1) +
  theme(text = element_text(family = "Lexend"),
        legend.position = "top",
        plot.title = element_text(size = 16, face = "bold",  hjust = 0.5)) +
  labs(fill = "Mean Percent Unobligated Funds",
       title = "Average Percent Unobligated TANF Funds by State")


usa <- map_data("usa")
state <- map_data("state")
ggplot(data=state, aes(x=long, y=lat, fill=region, group=group)) + 
  geom_polygon(color = "white", aes) + 
  guides(fill=FALSE) + 
  theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank(),
        axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) + 
  coord_fixed(1.3)

```

```{r}
#| warning: false
expenditures_bar3 <- expenditures |>
  group_by(state) |>
  summarize(mean_direct_assistance = mean(basic_assistance / total_expenditures)) |>
  ungroup() |>
  mutate(state = fct_rev(state))

us_hexbin2 <- st_read("data/us_states_hexgrid.geojson") |>
  mutate(google_name = toupper(str_remove(google_name, " \\(United States\\)"))) |>
  mutate(google_name = if_else(google_name == "DISTRICT OF COLUMBIA", "DIST.OF COLUMBIA", google_name)) |>
  left_join(expenditures_bar3, join_by(google_name == state))


ggplot(us_hexbin2) +
  geom_sf(aes(fill = mean_direct_assistance), color = "white", linewidth = .5) +
  geom_sf_text(aes(label = iso3166_2), color = "black" ) +
  theme_void() +
  scale_fill_distiller(direction = 1) +
  theme(legend.position = "top",
        plot.title = element_text(size = 16, face = "bold",  hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  labs(fill = "Mean Percent Direct Assistance")



```

```{r}
#| fig-height: 14
#| fig-width: 10


# wrangle data for bar plot
expenditures_bar4 <- expenditures |>
  filter(fiscal_year %in% c(2017:2021)) |>
  mutate(services = other_assistance + non_assistance + work_ed + early_ed +
           financial_ed + eitc_refundable + non_eitc_refundable +
           short_term_benefits + support__services + children_youth_services +
           pregnancy_prevention + fatherhood_programs + child_welfare +
           home_visiting + program_mgmt + other,
         total = services + basic_assistance + unobligated_balance,
         unob = unobligated_balance) |>
  select(state, fiscal_year, basic_assistance, services, unobligated_balance, total, unob) |>
  pivot_longer(cols = c(basic_assistance:unobligated_balance),
               names_to = "category",
               values_to = "amount")  |>
  group_by(state, category) |>
  summarize(amount = mean(amount),
            total = mean(total),
            unob = mean(unob),
            .groups = "drop") |>
  mutate(pct_unob = unob/total) |>
  mutate(state = fct_reorder(state, pct_unob))

expenditures_bar4 |>
  filter(state != "U.S. TOTAL") |>
  mutate(category = case_match(
    category,
    "basic_assistance" ~ "Basic Assistance",
    "services" ~ "Programs & Services",
    "unobligated_balance" ~ "Unobligated Balance"
  )) |>
  ggplot(aes(y = state)) +
  geom_col(aes(x = amount, fill = category), position = "fill") +
  geom_vline(xintercept = c(.25, .5, .75), color = "white", linetype = "dotted", linewidth = 1) + 
  scale_x_continuous(expand = c(0,0), labels = percent) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(legend.position = "top",
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        plot.background = element_rect(fill = "white", color = "white"),
        axis.text.y = element_text(color = "black"),
        panel.background = element_rect(fill = "white", color = "white"),
        panel.border = element_blank(),
        legend.key.size = unit(.25, 'cm')) +
  coord_cartesian(clip = "off") +
  guides(fill = guide_legend(title = NULL, reverse = T, override.aes = list(size = 4))) + 
  labs(x = "Average percent of total allocated funds (Combined Federal Funds and MOE)",
       y = NULL,
       fill = "Spending Category")

ggsave("unobligated_bar.png", width = 8, height = 10)

```


